### 2025-09-11 YEAR2/ECE253_LectureNotes/main.tex
911

\section{Information Representation}

\subsection{Number Systems}
\begin{definition}[Number System]
    A number system is a way of representing numbers using a set of symbols (digits) and a base (radix). The base determines the number of unique digits that can be used in the number system.
\end{definition}

@@

\section{Logic Circuits}
\begin{definition}[Logic Function]
    A logic function $L: \{0,1\}^n \rightarrow \{0,1\}$ is a mathematical function that takes $n$ binary inputs and produces a single binary output based on a set of rules.
\end{definition}

@@

>>>>

\section{Logic Circuits}
\subsection{Number Systems}
\begin{definition}[Number System]
    A number system is a way of representing numbers using a set of symbols (digits) and a base (radix). The base determines the number of unique digits that can be used in the number system.
\end{definition}

@@

\begin{definition}[Logic Function]
    A logic function $L: \{0,1\}^n \rightarrow \{0,1\}$ is a mathematical function that takes $n$ binary inputs and produces a single binary output based on a set of rules.
\end{definition}

@@

\subsection{Combinational Logic Circuits}

\subsection{Sum of Products (SOP) Form}
\begin{definition}[Literal]
    A literal is a variable or its negation. For example, $A$ and $\overline{A}$ are literals.  A literal can be either true or false, and it represents a single value in a logical expression.

\begin{definition}[Product Term]
    A product term is a logical synonym for AND.
\end{definition}

\begin{definition}[Sum Term]
    A sum term is a logical synonym for OR.
\end{definition}

\begin{definition}[Sum of Products (SOP) Form]
    A logical expression is in sum of products (SOP) form if it is a sum of product terms. For example, the expression $AB + \overline{A}C + BC$ is in SOP form.
\end{definition}

\begin{definition}[Minterm]
    A product term that evaluates to one for exactly one row of the truth table is called a minterm.
\end{definition}
\begin{example}[Minterm]
    For a given truth table for $x_1, x_2, x_3$, the minterms are:
    \begin{table}[h!]
        \centering
        \begin{tabular}{|c|c|c|c|}
            \hline
            $x_1$ & $x_2$ & $x_3$ & Minterm \\
            \hline
            0 & 0 & 0 & $m_0 = \overline{x_1}\overline{x_2}\overline{x_3}$ \\
            0 & 0 & 1 & $m_1 = \overline{x_1}\overline{x_2}x_3$ \\
            0 & 1 & 0 & $m_2 = \overline{x_1}x_2\overline{x_3}$ \\
            0 & 1 & 1 & $m_3 = \overline{x_1}x_2x_3$ \\
            1 & 0 & 0 & $m_4 = x_1\overline{x_2}\overline{x_3}$ \\
            1 & 0 & 1 & $m_5 = x_1\overline{x_2}x_3$ \\
            1 & 1 & 0 & $m_6 = x_1x_2\overline{x_3}$ \\
            1 & 1 & 1 & $m_7 = x_1x_2x_3$ \\
            \hline
        \end{tabular}
        \caption{Minterms for 3 Variables}
        \label{tab:minterms}

    \end{table}
    \\
    Note that each minterm corresponds to a unique combination of input values that produces an output of 1. To create the minterm, you would try to make every literal one.

\end{example}

\begin{definition}[Canonical SOP Form]
    A logical expression is in canonical SOP form if it is a sum of minterms. 
\end{definition}

\subsection{Product of Sums (POS) Form}

\begin{definition}[Product of Sums (POS) Form]
    A logical expression is in product of sums (POS) form if it is a product of sum terms. For example, the expression $(A + B)(\overline{A} + C)(B + C)$ is in POS form.
\end{definition}

\begin{definition}[Maxterm]
    A sum term that evaluates to zero for exactly one row of the truth table is called a maxterm.
\end{definition}

\begin{example}[Maxterm]
    For a given truth table for $x_1, x_2, x_3$, the maxterms are:
    \begin{table}[h!]
        \centering
        \begin{tabular}{|c|c|c|c|}
            \hline
            $x_1$ & $x_2$ & $x_3$ & Maxterm \\
            \hline
            0 & 0 & 0 & $M_0 = (x_1 + x_2 + x_3)$ \\
            0 & 0 & 1 & $M_1 = (x_1 + x_2 + \overline{x_3})$ \\
            0 & 1 & 0 & $M_2 = (x_1 + \overline{x_2} + x_3)$ \\
            0 & 1 & 1 & $M_3 = (x_1 + \overline{x_2} + \overline{x_3})$ \\
            1 & 0 & 0 & $M_4 = (\overline{x_1} + x_2 + x_3)$ \\
            1 & 0 & 1 & $M_5 = (\overline{x_1} + x_2 + \overline{x_3})$ \\
            1 & 1 & 0 & $M_6 = (\overline{x_1} + \overline{x_2} + x_3)$ \\
            1 & 1 & 1 & $M_7 = (\overline{x_1} + \overline{x_2} + \overline{x_3})$ \\
            \hline
        \end{tabular}
        \caption{Maxterms for 3 Variables}
        \label{tab:maxterms}
    \end{table}
    \\
    Note that each maxterm corresponds to a unique combination of input values that produces an output of 0. To create the maxterm, you would try to make every literal zero.
\end{example}

\begin{definition}[Canonical POS Form]
    A logical expression is in canonical POS form if it is a product of maxterms.
\end{definition}

\begin{theorem}[Converting between Canonical Forms]
    Any logical expression can be converted from canonical SOP form to canonical POS form and vice versa. For $i \in \{0, 1, \ldots, 2^n - 1\}$ and $S \subseteq \{0, 1, \ldots, 2^n - 1\}$, we have:
    \[
        f(x_1, x_2, \ldots, x_n) = \sum_{i \in S} m_i = \prod_{i \notin S} M_i
    \]
\end{theorem}
\begin{example}
    We have the following converison:
    $$
        f(x_1, x_2, x_3) = m_1 + m_3 + m_5 + m_7 = M_0 M_2 M_4 M_6
    $$
\end{example}

### 2025-09-11 YEAR2/PHY293_LectureNotes/main.tex
911

\begin{proof}
Let $A = \sqrt{A_1^2 + A_2^2}$ and choose $\phi$ such that 
\[
\cos(\phi) = \frac{A_1}{A}, \qquad \sin(\phi) = -\frac{A_2}{A}.
\]
Then, we can rewrite our original solution as
\begin{align*}
    x(t) &= A_1 \cos(\omega t) - A_2 \sin(\omega t) \\
         &= A \cos(\phi) \cos(\omega t) - A \sin(\phi) \sin(\omega t) \\
         &= A \big[\cos(\phi)\cos(\omega t) - \sin(\phi)\sin(\omega t)\big] \\ 
         &= A \cos(\omega t + \phi),
\end{align*}
by the cosine addition formula.
\end{proof}

@@

    To skip the math, lets assume a solution of the form \( x(t) = e^{i\omega t} \), substituting into the differential equation gives us a formulation for \( \omega \):
    \begin{align}
        -\omega^2 - i\gamma\omega + \omega_0^2 &= 0 \nonumber \\
        \omega &= -i\frac{\gamma}{2} \pm \sqrt{\omega_0^2 - \frac{\gamma^2}{4}} \label{eq:omega_damped}
    \end{align}
    Also, we can characterize the real and imaginary parts of \( \omega \) as:
    \begin{subequations}
        \begin{equation}
            \omega_r = \mathrm{Re}(\omega) = \pm \sqrt{\omega_0^2 - \frac{\gamma^2}{4}}
        \end{equation}
        \begin{equation}
            \omega_i = \mathrm{Im}(\omega) = -\frac{\gamma}{2}
        \end{equation}
    \end{subequations}
    The general solution for the damped harmonic oscillator can be written as:
    \begin{equation}
        x(t) = \exp(\omega_i t) \exp(-i \omega_r t) = \exp\left(-\frac{\gamma}{2}t\right) \exp(\mp i \sqrt{\omega_0^2 - \frac{\gamma^2}{4}} t)
    \end{equation}
    \begin{subequations}
    \begin{itemize}
        \item \textbf{No Damping} (\( \gamma = 0 \)): The system behaves like a simple harmonic oscillator with angular frequency \( \omega_0 \). Given by:
        \begin{equation}
            z = \exp(-i\omega_0 t)
        \end{equation}
        \item \textbf{Underdamping} (\( 0 < \gamma < 2\omega_0 \)): The system oscillates with a gradually decreasing amplitude. The angular frequency of oscillation is given by \( \omega_r = \sqrt{\omega_0^2 - \frac{\gamma^2}{4}} \). Given by:
        \begin{equation}
            z = \exp\left(-\frac{\gamma}{2}t\right) \exp(-i \omega_r t)
        \end{equation}
        The trigonometric form of the solution is:
        \begin{equation}
            x(t) = A_0 \exp\left(-\frac{\gamma}{2}t\right) \cos(\omega_r t + \phi)
        \end{equation}
        where \( A_0 \) and \( \phi \) are constants determined by the initial conditions
        From this, we can derive the following cases:
        \item \textbf{Critical Damping} (\( \gamma = 2\omega_0 \)): The system returns to equilibrium as quickly as possible without oscillating. The solution is given by:
        \begin{equation}
            z = \exp(-\omega_0 t)
        \end{equation}\
        \item \textbf{Overdamping} (\( \gamma > 2\omega_0 \)): The system returns to equilibrium without oscillating, but more slowly than in the critically damped case. The solution is given by:
        \begin{equation}
            z = \exp\left(-\frac{\gamma}{2}t\right) \exp\left(\sqrt{\frac{\gamma^2}{4} - \omega_0^2} t\right)
        \end{equation}
        So the general solution is (the solution is via a substitution of $x(t) = e^{-\gamma t/2} f(t)$ into the differential equation, which resolves the ODE to a simple form):
        \begin{equation}
            x(t) = A_1 \exp\left[\left(-\frac{\gamma}{2} + \sqrt{\frac{\gamma^2}{4} - \omega_0^2}\right) t\right] + A_2 \exp\left[\left(-\frac{\gamma}{2} - \sqrt{\frac{\gamma^2}{4} - \omega_0^2}\right) t\right]
        \end{equation}
        where \( A_1 \) and \( A_2 \) are constants determined by the initial conditions.
    \end{itemize}
    \end{subequations}
\end{definition}
\chapter{Modern Physics}
\end{document}

>>>>

\begin{proof}
Let $A = \sqrt{A_1^2 + A_2^2}$ and choose $\phi$ such that 
\[
\cos(\phi) = \frac{A_1}{A}, \qquad \sin(\phi) = -\frac{A_2}{A}.
\]
Then, we can rewrite our original solution as
\begin{align*}
    x(t) &= A_1 \cos(\omega t) + A_2 \sin(\omega t) \\
         &= A \cos(\phi) \cos(\omega t) - A \sin(\phi) \sin(\omega t) \\
         &= A \big[\cos(\phi)\cos(\omega t) - \sin(\phi)\sin(\omega t)\big] \\ 
         &= A \cos(\omega t + \phi),
\end{align*}
by the cosine addition formula.
\end{proof}

@@

    To skip the math, lets assume a solution of the form \( x(t) = e^{i\omega t} \), substituting into the differential equation gives us a formulation for \( \omega \):
    \begin{align}
        -\omega^2 - i\gamma\omega + \omega_0^2 &= 0 \nonumber \\
        \omega &= -i\frac{\gamma}{2} \pm \sqrt{\omega_0^2 - \frac{\gamma^2}{4}} \label{eq:omega_damped}
    \end{align}
    Also, we can characterize the real and imaginary parts of \( \omega \) as:
    \begin{subequations}
        \begin{equation}
            \omega_r = \mathrm{Re}(\omega) = \pm \sqrt{\omega_0^2 - \frac{\gamma^2}{4}}
        \end{equation}
        \begin{equation}
            \omega_i = \mathrm{Im}(\omega) = -\frac{\gamma}{2}
        \end{equation}
    \end{subequations}
    The general solution for the damped harmonic oscillator can be written as:
    \begin{equation}
        x(t) = \exp(\omega_i t) \exp(-i \omega_r t) = \exp\left(-\frac{\gamma}{2}t\right) \exp(\mp i \sqrt{\omega_0^2 - \frac{\gamma^2}{4}} t)
    \end{equation}
    \begin{subequations}
    \begin{itemize}
        \item \textbf{No Damping} (\( \gamma = 0 \)): The system behaves like a simple harmonic oscillator with angular frequency \( \omega_0 \). Given by:
        \begin{equation}
            z = \exp(-i\omega_0 t)
        \end{equation}
        \item \textbf{Underdamping} (\( 0 < \gamma < 2\omega_0 \)): The system oscillates with a gradually decreasing amplitude. The angular frequency of oscillation is given by \( \omega_r = \sqrt{\omega_0^2 - \frac{\gamma^2}{4}} \). Given by:
        \begin{equation}
            z = \exp\left(-\frac{\gamma}{2}t\right) \exp(-i \omega_r t)
        \end{equation}
        The trigonometric form of the solution is:
        \begin{equation}
            x(t) = A_0 \exp\left(-\frac{\gamma}{2}t\right) \cos(\omega_r t + \phi)
        \end{equation}
        where \( A_0 \) and \( \phi \) are constants determined by the initial conditions
        From this, we can derive the following cases:
        \item \textbf{Critical Damping} (\( \gamma = 2\omega_0 \)): The system returns to equilibrium as quickly as possible without oscillating. Consider:
        $$
            x(t) = e^{-\frac{\gamma}{2}t}f(t)
        $$
        Inserting into the differential equation, we get:
        $$
            \ddot{f} + \left(\omega_0^2 - \frac{\gamma^2}{4}\right)f = 0
        $$
        Since \( \gamma = 2\omega_0 \), we have \( \omega_0^2 - \frac{\gamma^2}{4} = 0 \), leading to:
        $$
            \ddot{f} = 0 \implies f(t) = A_1 t + A_2
        $$
        Therefore, the general solution for the critically damped case is:
        \begin{equation}
            x(t) = (A_1 t + A_2) \exp\left(-\frac{\gamma}{2}t\right)
        \end{equation}
        where \( A_1 \) and \( A_2 \) are constants determined by
        \item \textbf{Overdamping} (\( \gamma > 2\omega_0 \)): The system returns to equilibrium without oscillating, but more slowly than in the critically damped case. The solution is given by:
        \begin{equation}
            z = \exp\left(-\frac{\gamma}{2}t\right) \exp\left(\sqrt{\frac{\gamma^2}{4} - \omega_0^2} t\right)
        \end{equation}
        So the general solution is (the solution is via a substitution of $x(t) = e^{-\gamma t/2} f(t)$ into the differential equation, which resolves the ODE to a simple form):
        \begin{equation}
            x(t) = A_1 \exp\left[\left(-\frac{\gamma}{2} + \sqrt{\frac{\gamma^2}{4} - \omega_0^2}\right) t\right] + A_2 \exp\left[\left(-\frac{\gamma}{2} - \sqrt{\frac{\gamma^2}{4} - \omega_0^2}\right) t\right]
        \end{equation}
        where \( A_1 \) and \( A_2 \) are constants determined by the initial conditions.
    \end{itemize}
    \end{subequations}
\end{definition}

\subsection{Energy and Quality Factor}
\begin{definition}[Energy of a Very Light Damping]
    Consider a very lightly damped harmonic oscillator, where \( \gamma \ll \omega_0 \). In this case, the angular frequency of oscillation \( \omega_r \) can be approximated as:
    $$
    \omega_r \approx \omega_0 \left(1 - \frac{\gamma^2}{8\omega_0^2}\right) \approx \omega_0
    $$
    So the motion of the lightly damped oscillator can be approximated as:
    $$
    x(t) \approx A_0 e^{-\frac{\gamma}{2}t} \cos(\omega_0 t + \phi)
    $$
    Then, we can calculate the velocity of the oscillator:
    \begin{align*}
        \dot{x}(t) &= -\frac{\gamma}{2} A_0 e^{-\frac{\gamma}{2}t} \cos(\omega_0 t + \phi) - A_0 \omega_0 e^{-\frac{\gamma}{2}t} \sin(\omega_0 t + \phi) \\
        &= A_0 \omega_0 e^{-\frac{\gamma}{2}t} \left(-\frac{\gamma}{2\omega_0} \cos(\omega_0 t + \phi) - \sin(\omega_0 t + \phi)\right)
    \end{align*}
    The total mechanical energy \( E(t) \) of the lightly damped oscillator is given by:
    \begin{align*}
        E(t) &= \frac{1}{2}m\dot{x}^2 + \frac{1}{2}kx^2 \\
             &= \frac{1}{2}m \left[A_0 \omega_0 e^{-\frac{\gamma}{2}t} \left(-\frac{\gamma}{2\omega_0} \cos(\omega_0 t + \phi) - \sin(\omega_0 t + \phi)\right)\right]^2 + \frac{1}{2}k \left[A_0 e^{-\frac{\gamma}{2}t} \cos(\omega_0 t + \phi)\right]^2 \\
             &= \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \left[\left(-\frac{\gamma}{2\omega_0} \cos(\omega_0 t + \phi) - \sin(\omega_0 t + \phi)\right)^2 + \cos^2(\omega_0 t + \phi)\right] \\
             &= \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \left[\sin^2(\omega_0 t + \phi) + \cos^2(\omega_0 t + \phi) + \frac{\gamma^2}{4\omega_0^2} \cos^2(\omega_0 t + \phi) + \frac{\gamma}{\omega_0} \sin(\omega_0 t + \phi) \cos(\omega_0 t + \phi)\right] \\
             &\approx \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \left[1 + \frac{\gamma^2}{4\omega_0^2} \cos^2(\omega_0 t + \phi)\right] \quad (\text{neglecting the small term } \frac{\gamma}{\omega_0} \sin(\omega_0 t + \phi) \cos(\omega_0 t + \phi)) \\
             &\approx \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \quad (\text{since } \frac{\gamma^2}{4\omega_0^2} \text{ is very small}) \\
             &= E_0 e^{-\gamma t} \quad \text{where } E_0 = \frac{1}{2}m A_0^2 \omega_0^2 \text{ is the initial energy at } t=0
    \end{align*}
    We can also define the time constant \( \tau \) as the time it takes for the energy to decrease to \( \frac{1}{e} \) of its initial value:
    \begin{equation}
        \tau = \frac{1}{\gamma}
    \end{equation}
    So we have, for very light damping:
    \begin{equation}
        E(t) = E_0 e^{-\frac{t}{\tau}}
    \end{equation}
\end{definition}

\begin{definition}[Rate of Energy Loss]
    Taking the time derivative of the total mechanical energy \( E(t) \):
    \begin{align*}
        \frac{dE}{dt} &= \frac{d}{dt} \left(\frac{1}{2}m\dot{x}^2 + \frac{1}{2}kx^2 \right) \\
                      &= (ma + kx)\dot{x} \\
    \end{align*}
    For a undamped harmonic oscillator, \( ma + kx = 0 \), so \( \frac{dE}{dt} = 0 \), indicating that the total mechanical energy is conserved and obeys Hooke's law completely. However, for a damped harmonic oscillator, \( ma + kx = -b\dot{x} \), leading to:
    \begin{equation}
        \frac{dE}{dt} = -b\dot{x}^2
    \end{equation}
\end{definition}

\begin{definition}[Quality Factor (Q-Factor)]
    The quality factor \( Q \) is a dimensionless parameter that characterizes the damping of a harmonic oscillator. It is defined as:
    \begin{subequations}
    \begin{equation}
        Q = \frac{\omega}{\gamma} = \omega \tau
    \end{equation}
    And for very light damping, we can approximate \( \omega \approx \omega_0 \), leading to:
    \begin{equation}
        Q \approx \frac{\omega_0}{\gamma} = \omega_0 \tau
    \end{equation}
    \end{subequations}
    This allows us to rewrite the equation of a damped harmonic oscillator as:
    \begin{equation}
        \ddot{x} + \frac{\omega_0}{Q} \dot{x} + \omega_0^2 x = 0
    \end{equation}
    and:
    \begin{equation}
        \omega = \omega_0 \sqrt{1 - \frac{1}{4Q^2}}
    \end{equation}
\end{definition}
\chapter{Modern Physics}
\end{document}
